Anomaly detection with kernel density estimation on manifolds

### 2022-01-18 paper/paper.Rmd
Paper setup

---
title: "Some ideas about anomalies on manifolds"
author: "Anastasios Panagiotelis"
date: "05/02/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dimRed)
library(rgl)
library(hdrcde)
knitr::knit_hooks$set(webgl = hook_webgl)
set.seed(1983)
```

## Background

Consider the case of points on a unit-radius sphere with three coordinates $x$, $y$ and $z$.  Consider only the positive quadrant, i.e. points where $x>0$,$y>0$,$z>0$.  This is not just a toy example, for instance consider a statistical manifold for discrete distributions of a random variable $U$ with domain 0,1,2.  If we set $x=\sqrt{\textrm{Pr}(U=0)}$, $y=\sqrt{\textrm{Pr}(U=1)}$, $z=\sqrt{\textrm{Pr}(U=2)}$ then each distribution is a point that lies in the positive quadrant of unit sphere.

There are three coordinates since the ambient space is $\mathbb{R}^3$ but the surface of the sphere itself is 2-dimensional.  There are a number of ways we can reduce the dimension down to two coordinates.  For instance we can use the polar coordinates given by

$$
\theta=arccos(z)\\
\phi=arctan(y/x)
$$

Here, $\theta$ is an angle downwards from the north pole and $\phi$ is an angle around the equator.  For our purposes, think of computing polar coordinates as a way of doing dimension reduction, i.e. you input $(x,y,z)$ and it returns $(\theta,\phi)$.  Although an actual manifold learning algorithm is not going to return the polar coordinates, it serves as a nice illustration of some of the issues around identifying anomalies on manifolds.

## Uniform in sphere $\neq$ uniform in polar coordinates

For a more extensive discussion of parts of this section see [this blog post](http://corysimon.github.io/articles/uniformdistn-on-sphere/) noting that they use $\phi$ and $\theta$ the other way around.  Suppose we want to generate data uniformly on the sphere.  Without getting into too much detail as to why, the following will work:

```{r unifsphere, echo=TRUE, eval=TRUE}
n<-1000 #number of observations to be simulated

#Generate from independent standard normals with constant variance (the fact that constant variance and zero covariance is called sphericity is not a coincidence)

x<-rnorm(n)
y<-rnorm(n)
z<-rnorm(n)

#Norm of randomly generated vectors
norm_xyz<-sqrt(x^2+y^2+z^2)

#Normalise each coordinate to guarantee they lie on unit sphere
#Taking the absolute value puts everything into the upper quadrant
x<-abs(x)/norm_xyz 
y<-abs(y)/norm_xyz
z<-abs(z)/norm_xyz

```

This can be plotted in three dimensions and looked at interactively by clicking and dragging

```{r unisphereplot, webgl=TRUE, echo=TRUE}

plot3d(x,y,z)
```

Suppose we want to understand something about the density of these points.  Strictly speaking anomaly detection (Robs colorful plots) will break down a bit because these are uniform data and there are no anomalies.  However the following exercise does give intuition.

Suppose our 'manifold learning' algorithm is to apply the functions that give us spherical coordinates

```{r sphcoord, echo=TRUE, eval=TRUE}
theta<-acos(z)
phi<-atan(y/x)
```

below is a plot of the spherical coordinates with HDRs.

```{r hdeplot}
plot(hdrscatterplot(theta,phi))
```

Notice that the region with low values of $\theta$ is sparse and the data is not uniform.  This is the region near the north pole that gets stretched out when we transform from three dimensional coordinates to polar coordinates.  The anomalies we have detected are a consequence of the embedding and NOT a feature of the data.  More generally looking at the distribution of the low-dimensional data does not always tell us the correct things about the distribution of the high dimensional data

>>>>

---
title: "Anomaly detection with kernel density estimation on manifolds"
author:
- familyname: Cheng
  othernames: Fan
  address: Monash University
  email: Fan.Cheng@monash.edu
  # correspondingauthor: true
- familyname: Hyndman
  othernames: Rob J
  address: Monash University
  email: Rob.Hyndman@monash.edu
- familyname: Panagiotelis
  othernames: Anastasios
  address: University of Sydney
  email: Anastasios.Panagiotelis@sydney.edu.au
abstract: ""
# keywords: "manifold learning"
# wpnumber: 03/2021
# jelcodes: C55, C65, C80
blind: true
cover: false
toc: false
lot: false
bibliography: references.bib
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
      # after_body: appendix.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
    latex_engine: pdflatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = FALSE,
  messages = FALSE,
  warning = FALSE,
  eval = TRUE
  # include = FALSE
)
options(tinytex.verbose = TRUE)
library(dimRed)
library(tidyverse)
library(ggplot2)
library(reshape2)
library(knitr)
library(kableExtra)
library(tibble)
library(patchwork)
library(data.table)
library(dtplyr)
library(igraph)
library(ggraph)
# set.seed(1)
Jmisc::sourceAll(here::here("R/sources"))
```

# Introduction
